Stochastic Variance-Reduced Policy Gradient

### 2018-02-07 svrpg.tex
Updated

\begin{algorithm}[tb]
	\caption{SVRG}
	\label{alg:svrg}
	\begin{algorithmic}
            \STATE {\bfseries Input:} a dataset $\mathcal{D}_N$, number of epochs $S$, epoch size $m$, step size $\alpha$, initial parameter $\vtheta_{0}^0:=\wt{\vtheta}^0$
		\FOR{$s=0$ {\bfseries to} $S-1$}
        \STATE $\wt{\mu} = \nabla f(\wt{\vtheta}^s)$
		\FOR{$t=0$ {\bfseries to} $m-1$}
%		\STATE Pick $i$ uniformly at random from $[1,N]$
        \STATE $x_i \sim \mathcal{U}\left(\mathcal{D}_N\right)$
		\STATE $v^{s}_t = 
%			\nabla f(\wt{\vtheta}^s) + 
			\wt{\mu} + 
			\nabla g(x_i|\vtheta_t^{s}) -
			\nabla g(x_i|\wt{\vtheta}^{s})
		$
        \STATE $\vtheta_{t+1}^{s} = \vtheta_t^{s} + \alpha v^{s}_t$
		\ENDFOR
        \STATE $\vtheta_0^{s+1} := \wt{\vtheta}^{s+1} = \vtheta_{m}^s$
		\ENDFOR
        \STATE \underline{Concave case:} \textbf{return} $\vtheta_{0}^S$
        \STATE \underline{Non-Concave case:} \textbf{return} $\vtheta_t^s$ with $(s,t)$ picked uniformly at random from $\{[0,S[\times[0,m[\} \cup \{(S,0)\}$
	\end{algorithmic}
\end{algorithm}
\todomat{Mention the fact that we return last parameter in practice}

@@

\textbf{Adaptive step size.}
A common way to deal with noisy gradients is to use adaptive strategies to compute the step size.
ADAptive Moment estimation (ADAM)~\citep{kingma2014adam} stabilizes the parameter update by computing learning rates for each parameter based on an incremental estimate of the gradient variance.
Due to this feature we would like to incorporate ADAM in the structure of the SVRPG update.
Recall that SVRPG is characterized by two updates: I) FG update in the snapshot; II) corrected gradient update in the sub-iterations.
Given this structure, we suggest to use two ADAM estimators:
\begin{align*}
        \vtheta^{s}_1 &= \wt{\vtheta}^s + \alpha^{\textsc{FG}}_s\wh{\nabla}_N J(\wt{\vtheta}^s)\\
        \vtheta^{s}_{t+1} &= \vtheta^{s}_t + \alpha^{\textsc{SI}}_{s,t} v_t^{s}
\end{align*}
where $\alpha^{\textsc{FG}}_{s}$ is associated to the snapshot and $\alpha^{\textsc{SI}}_{s,t}$ with the sub-iterations (see App.~\ref{app:practicalsvrpg} for details).
By doing so we decouple the contribution of the variance introduced by the approximate FG from the one introduced by the sub-iterations.
Note that these two terms have different orders of magnitude since are estimated with a different number of trajectories ($B \ll N$) and w/o importance weights.
The use of two ADAM estimators allows to capture and exploit this property.

@@

\subsection{Normalized Importance Sampling}\label{sec:prac}
As we already mentioned in section \ref{sec:stopping} importance weights means a lot of variance to deal with. Therefore, in order to cope with this issue the literature propose the Self-normalization techniques.
This technique offers a variance reduction over the importance weights in exchange of a bias introduction. Hence, if the bias introduced by this is smaller than the variance reduction it causes, it is convenient to use it in practice.  
As regards our algorithm, the term
$
 \frac{1}{B} \sum_{i=0}^{B-1} \left[ \omega(\tau_i|\vtheta_t, \wt{\vtheta}) g(\tau_i|\wt{\vtheta})
\right]
$  \eqref{E:svrpg.estimate.batch} become:
\begin{align*}
\frac{1}{ \sum_{i=0}^{B-1}\omega(\tau_i|\vtheta_t, \wt{\vtheta})} \sum_{i=0}^{B-1} \left[ \omega(\tau_i|\vtheta_t, \wt{\vtheta}) g(\tau_i|\wt{\vtheta})
\right].
\end{align*}
The rest of terms remain unchanged.
We experienced some improvements using it (see section \ref{sec:exp} for the experimental point of view), in particular in our experiments we use this technique in the context of G(PO)MDP defined as "weighted per-decision importance sampling estimator" \cite{precup2000eligibility}.

\subsection{Baseline}
As reported in Sec. \ref{subsec:PolicyGradient} a known method for variance reduction in Reinforcement Learning is the use of baseline.
This method is independent to SVRPO.
We decide to test SVRPO with an Actor Critic baseline: the baseline is a states dependent linear function with a time-varying feature vector \cite{duan2016benchmarking}.
Our choice is to use a single baseline for each term of expression \ref{E:svrpg.estimate.batch}. This is update only in the snapshot: the baseline is estimated using the N trajectories.

@@

Concerning importance weighting techniques, RL has made wide use of them for off-policy problems~\citep[\eg]{precup2000eligibility,thomas2015high}. However, as mentioned before, SVRPG cannot be compared to such methods since it is practically on-policy. Here, importance weighting is just a statistical tool used to preserve the unbiasedness of the corrected gradient.

@@

\textbf{On-policy setting.}
Consider a policy $\pol$ and let $\mathcal{D}_N = \{\tau_i\}_{i=1}^N$ be a dataset collected using policy $\pol$.
The REINFORCE gradient estimator~\citep{williams1992simple} provides a simple, unbiased way of estimating the gradient:
\begin{align*}
\gradApp{\vtheta}{N} = \frac{1}{N}\sum_{n=1}^{N}
\underbracket{
\left(\sum_{h=0}^{H}\score{\vtheta}{a_h^n\vert s_h^n}\right)\left(\sum_{h=0}^{H}\gamma^h r(s_h^n,a_h^n) - b\right)
}_{g(\tau_n|\vtheta)}
,
\end{align*}
where subscripts denote the time step, superscripts denote the trajectory, $r_h$ is the reward actually collected at time $h$ and b can be any baseline provided $\EVV[a \sim \pol]{b(s,a)} = 0$, for any $s\in\Sspace$.\todopir{check definition of baseline}
The G(PO)MDP gradient estimator~\cite{baxter2001infinite} is a refinement of REINFORCE which is subject to less variance \cite{zhao2011analysis} while preserving the unbiasedness:
\begin{align*}
\gradApp{\vtheta}{N} = \frac{1}{N}\sum_{n=1}^{N}
\underbracket{
\sum_{h=0}^{H}\left(\sum_{k=0}^{h}\score{\vtheta}{a_h^n\vert s_h^n}\right)\left(\gamma^h r(s_h^n,a_h^n) - b\right)
}_{g(\tau_n|\vtheta)}.
\end{align*}
G(PO)MDP can be seen as a more efficient implementation
of the REINFORCE algorithm. In fact the latter does not perform
an optimal credit assignment, since it ignores that the reward at time $t$ does
not depend on the action performed after time $t$. G(PO)MDP overcomes this
issue taking into account the causality of rewards in the REINFORCE definition
of policy gradient.

\textbf{Off-Policy setting.}
In off-policy setting two policies, called behavioural $\pi^B$ and target $\pi^T$, are involved.
The first is used to select actions for the interaction with the system, while the second is used to evaluate the agent performance and it is improved in each update.
Suppose now that we aim to estimate the performance of the target policy $\pi^T$ but we have samples collected using policy $\pi^B$.
We can use importance weight correction to correct the shift in the distribution and obtain an unbiased estimate of $J(\pi^T)$:
\[
        J(\pi^T) = \EVV[\tau \sim \Delta(\cdot|\pi^T)]{\mathcal{R}(\tau)} = \EVV[\tau \sim \Delta(\cdot|\pi^B)]{\omega(\tau) \mathcal{R}(\tau)}
\]
where $\omega(\tau) = \frac{\Delta(\tau,\pi^T)}{\Delta(\tau|\pi^B)} = \omega(z_{0:H}) = \prod_{w=0}^{H} \omega(z_w)$ and $\omega(z_w) = \frac{\pi^T(a_w|s_w)}{\pi^B(a_w|s_w)}$

The definition of the off-policy version of~\eqref{E:onpolicygradient} is~\citep[\eg][]{jurvcivcek2012reinforcement}
\begin{equation}\label{E:offpolicygradient}
        \nabla J(\pi^T) = \EVV[\tau \sim \Delta(\cdot|\pi^B)]{\omega(\tau) \nabla \log \Delta(\tau|\pi^T) \mathcal{R}(\tau)}
\end{equation}
For $\omega(\tau)$ being well defined the behavioural policy should have non-zero probability of selecting any action in every state \ie $\pi^B(a|s) > 0$ for any $(s,a)\in \mathcal{S} \times \mathcal{A}$.

@@

\section{Proofs}\label{app:proofs}
In this Appendix we prove all the claims made in the paper, with the primary objective of proving Theorem \ref{theo:convergence}. Our proof is adapted from the one of Theorem 2 from \cite{reddi2016stochastic} and has a very similar structure, but with all the additional challenges and assumptions described in Section \ref{sec:conv}.

@@

\begin{definition}
For a random variable $X$:
\begin{align*}
	&\mathbb{E}_{t\vert s}\left[X\right] \coloneqq 
		\mathop{\mathbb{E}}_{\substack{\tau_j\sim\pi(\cdot\vert\tilde{\vtheta}^s)\forall j \in {\it {\bf N}} \\ \tau_{i,h}\sim\pi(\cdot\vert\vtheta^{s+1}_i) \forall h \in {\it {\bf B}}, \text{ for $i=0,\dots,t-1$}}}{\left[X \vert \tilde{\vtheta^s}\right]} \\
	&\coloneqq \EVV[\tau_j\sim\pi(\cdot\vert\tilde{\vtheta}^s)\forall j \in {\it {\bf N}}]{
			\EVV[\tau_{0,h}\sim\pi(\cdot\vert\vtheta_0^{s+1}) \forall h \in {\it {\bf B}}]
				{\dots
					\EVV[\tau_{t,h}\sim\pi(\cdot\vert\vtheta_t^{s+1})\forall h \in {\it {\bf B}}]
						{X\vert\vtheta_t^{s+1}}
				 \dots
			\vert\vtheta_0^{s+1}}
		\vert\tilde{\vtheta}^s},
\end{align*}
where the sequence $\tilde{\vtheta}^s,\vtheta_0^{s+1},\dots,\vtheta_t^{s+1}$ is defined in Algorithm \ref{alg:svrpg}, ${\it {\bf N}} = [0,\dots,N)$, and ${\it {\bf B}} = [0,\dots,B)$. 
\end{definition}
Intuitively, the $\Ets{\cdot}$ operator computes the expected value with respect to the sampling of trajectories from the snapshot $\tilde{\vtheta}^s$ up to the $t$-th iteration included. Note that the order in which expected values are taken is important since each $\vtheta_{t}^{s+1}$ is function of previously sampled trajectories and is used to sample new ones.

>>>>

\begin{algorithm}[tb]
	\caption{SVRG}
	\label{alg:svrg}
	\begin{algorithmic}
            \STATE {\bfseries Input:} a dataset $\mathcal{D}_N$, number of epochs $S$, epoch size $m$, step size $\alpha$, initial parameter $\vtheta_{0}^0:=\wt{\vtheta}^0$
		\FOR{$s=0$ {\bfseries to} $S-1$}
        \STATE $\wt{\mu} = \nabla f(\wt{\vtheta}^s)$
		\FOR{$t=0$ {\bfseries to} $m-1$}
%		\STATE Pick $i$ uniformly at random from $[1,N]$
        \STATE $x_i \sim \mathcal{U}\left(\mathcal{D}_N\right)$
		\STATE $v^{s}_t = 
%			\nabla f(\wt{\vtheta}^s) + 
			\wt{\mu} + 
			\nabla z(x_i|\vtheta_t^{s}) -
			\nabla z(x_i|\wt{\vtheta}^{s})
		$
        \STATE $\vtheta_{t+1}^{s} = \vtheta_t^{s} + \alpha v^{s}_t$
		\ENDFOR
        \STATE $\vtheta_0^{s+1} := \wt{\vtheta}^{s+1} = \vtheta_{m}^s$
		\ENDFOR
        \STATE \underline{Concave case:} \textbf{return} $\vtheta_{0}^S$
        \STATE \underline{Non-Concave case:} \textbf{return} $\vtheta_t^s$ with $(s,t)$ picked uniformly at random from $\{[0,S[\times[0,m[\} \cup \{(S,0)\}$
	\end{algorithmic}
\end{algorithm}
\todomat{Mention the fact that we return last parameter in practice}

@@

\textbf{Adaptive step size.}
A common way to deal with noisy gradients is to use adaptive strategies to compute the step size.
ADAptive Moment estimation (ADAM)~\citep{kingma2014adam} stabilizes the parameter update by computing learning rates for each parameter based on an incremental estimate of the gradient variance.
Due to this feature we would like to incorporate ADAM in the structure of the SVRPG update.
Recall that SVRPG is characterized by two updates: I) FG update in the snapshot; II) corrected gradient update in the sub-iterations.
Given this structure, we suggest to use two ADAM estimators:
\begin{align*}
        \vtheta^{s}_1 &= \wt{\vtheta}^s + \alpha^{\textsc{FG}}_s\wh{\nabla}_N J(\wt{\vtheta}^s)\\
        \vtheta^{s}_{t+1} &= \vtheta^{s}_t + \alpha^{\textsc{SI}}_{s,t} v_t^{s}
\end{align*}
where $\alpha^{\textsc{FG}}_{s}$ is associated to the snapshot and $\alpha^{\textsc{SI}}_{s,t}$ with the sub-iterations (see App.~\ref{app:practicalsvrpg} for details).
\todopir{Add ADAM definition and explain in more details}
By doing so we decouple the contribution of the variance introduced by the approximate FG from the one introduced by the sub-iterations.
Note that these two terms have different orders of magnitude since are estimated with a different number of trajectories ($B \ll N$) and w/o importance weights.
The use of two ADAM estimators allows to capture and exploit this property.

@@

\subsection{Normalized Importance Sampling}\label{sec:prac}
As we already mentioned in section \ref{sec:stopping} importance weights means a lot of variance to deal with. Therefore, in oreder to cope with this issue the literature propose the Self-normalization techniques \cite{precup2000eligibility}.
This technique offers a variance reduction over the importance weights in exchange of a bias introduction. Hence, if the bias introduced by this is smaller than the variance reduction it causes, it is convenient to use it in practice.  We expirienced some improvements using it (see section \ref{sec:exp} for the sperimental point of view).
Self-normalization is easy to use, but depends on the chosen reinforcement learning algorithm. Therefore, here we give the definition of this technique in the context of G(PO)MDP where we have used the weighted per-decision importance sampling estimator \cite{precup2000eligibility}:
\begin{align*}
\gradApp{\vtheta}{N} = \frac{1}{\Omega}\sum_{n=1}^{N}\sum_{h=0}^{H-1}\left(\sum_{k=0}^{h}\score{\vtheta}{a_k^n\vert s_k^n}\right)\left(\gamma^h r_h^n \omega(z_{0:h})\right).
\end{align*} 
where:
\begin{align*}
\omega(z_{0:h}) = \prod_{k=0}^{h} \frac{\pi_{\vtheta}({a_k^n\vert s_k^n})
 }{\pi_{\vtheta^{'}}({a_k^n\vert s_k^n})}
\end{align*}
\begin{align*}
\Omega = \sum_{n=1}^{N}\sum_{h=0}^{H-1}\gamma^h\prod_{k=0}^{h} \frac{\pi_{\vtheta}({a_k^n\vert s_k^n})
}{\pi_{\vtheta^{'}}({a_k^n\vert s_k^n})}
\end{align*}
\todomat{why should we use it? What advantages do we expect? At what cost? Pointer to experiments}

@@

Concerning importance weighting techniques, RL has made wide use of them for off-policy problems~\citep[\eg][]{precup2000eligibility,thomas2015high}. However, as mentioned before, SVRPG cannot be compared to such methods since it is practically on-policy. Here, importance weighting is just a statistical tool used to preserve the unbiasedness of the corrected gradient.

@@

\textbf{On-policy setting.}
Consider a policy $\pol$ and let $\mathcal{D}_N = \{\tau_i\}_{i=1}^N$ be a dataset collected using policy $\pol$.
The REINFORCE gradient estimator~\citep{williams1992simple} provides a simple, unbiased way of estimating the gradient:
\begin{align*}
\gradApp{\vtheta}{N} = \frac{1}{N}\sum_{n=1}^{N}
\underbracket{
\left(\sum_{h=0}^{H}\score{\vtheta}{a_h^n\vert s_h^n}\right)\left(\sum_{h=0}^{H}\gamma^h r(s_h^n,a_h^n) - b\right)
}_{g(\tau_n|\vtheta):=\nabla \log\Delta(\tau_n|\vtheta)\Reward(\tau_n)}
,
\end{align*}
where subscripts denote the time step, superscripts denote the trajectory, $r_h$ is the reward actually collected at time $h$ and b can be any baseline provided $\EVV[a \sim \pol]{b(s,a)} = 0$, for any $s\in\Sspace$.\todopir{check definition of baseline}
The G(PO)MDP gradient estimator~\cite{baxter2001infinite} is a refinement of REINFORCE which is subject to less variance \cite{zhao2011analysis} while preserving the unbiasedness:
\begin{align*}
\gradApp{\vtheta}{N} = \frac{1}{N}\sum_{n=1}^{N}
\underbracket{
\sum_{h=0}^{H}\left(\sum_{k=0}^{h}\score{\vtheta}{a_h^n\vert s_h^n}\right)\left(\gamma^h r(s_h^n,a_h^n) - b\right)
}_{g(\tau_n|\vtheta)}.
\end{align*}
G(PO)MDP can be seen as a more efficient implementation
of the REINFORCE algorithm. In fact the latter does not perform
an optimal credit assignment, since it ignores that the reward at time $t$ does
not depend on the action performed after time $t$. G(PO)MDP overcomes this
issue taking into account the causality of rewards in the REINFORCE definition
of policy gradient.

\textbf{Off-policy setting.}
In off-policy setting two policies, called behavioural $\pi^B$ and target $\pi^T$, are involved.
The first is used to select actions for the interaction with the system, while the second is used to evaluate the agent performance and it is improved in each update.
Suppose now that we aim to estimate the performance of the target policy $\pi^T$ but we have samples collected using policy $\pi^B$.
We can use importance weight correction to correct the shift in the distribution and obtain an unbiased estimate of $J(\pi^T)$:
\[
        J(\pi^T) = \EVV[\tau \sim \Delta(\cdot|\pi^T)]{\mathcal{R}(\tau)} = \EVV[\tau \sim \Delta(\cdot|\pi^B)]{\omega(\tau) \mathcal{R}(\tau)}
\]
where $\omega(\tau) = \frac{\Delta(\tau,\pi^T)}{\Delta(\tau|\pi^B)} = \omega(z_{0:H}) = \prod_{w=0}^{H} \omega(z_w)$ and $\omega(z_w) = \frac{\pi^T(a_w|s_w)}{\pi^B(a_w|s_w)}$

The definition of the off-policy version of~\eqref{E:onpolicygradient} is~\citep[\eg][]{jurvcivcek2012reinforcement}
\begin{equation}\label{E:offpolicygradient}
        \nabla J(\pi^T) = \EVV[\tau \sim \Delta(\cdot|\pi^B)]{\omega(\tau) \nabla \log \Delta(\tau|\pi^T) \mathcal{R}(\tau)} = \EVV[\tau \sim \Delta(\cdot|\pi^B)]{\omega(\tau) g(\tau|\vtheta)}. 
\end{equation}
For $\omega(\tau)$ being well defined the behavioural policy should have non-zero probability of selecting any action in every state \ie $\pi^B(a|s) > 0$ for any $(s,a)\in \mathcal{S} \times \mathcal{A}$.
Equation~\ref{E:offpolicygradient} is important for proving Theorem~\ref{theo:convergence} since it provides a common representation of REINFORCE and G(PO)MDP.

@@

\section{Proofs}\label{app:proofs}
In this section, we prove all the claims made in the paper, with the primary objective of proving Theorem \ref{theo:convergence}.
Our proof is adapted from the one of Theorem 2 from \cite{reddi2016stochastic} and has a very similar structure, but with all the additional challenges and assumptions described in Section \ref{sec:conv}.

@@

\begin{definition}
For a random variable $X$:
\begin{align*}
        \mathbb{E}_{t\vert s}\left[X\right] &\coloneqq 
		\mathop{\mathbb{E}}_{\substack{\tau_j\sim\pi(\cdot\vert\tilde{\vtheta}^s)\forall j \in {\it {\bf N}} \\ \tau_{i,h}\sim\pi(\cdot\vert\vtheta^{s+1}_i) \forall h \in {\it {\bf B}}, \text{ for $i=0,\dots,t-1$}}}{\left[X \vert \tilde{\vtheta^s}\right]} \\
	&\coloneqq \EVV[\tau_j\sim\pi(\cdot\vert\tilde{\vtheta}^s)\forall j \in {\it {\bf N}}]{
			\EVV[\tau_{0,h}\sim\pi(\cdot\vert\vtheta_0^{s+1}) \forall h \in {\it {\bf B}}]
				{\dots
					\EVV[\tau_{t,h}\sim\pi(\cdot\vert\vtheta_t^{s+1})\forall h \in {\it {\bf B}}]
						{X\vert\vtheta_t^{s+1}}
				 \dots
			\vert\vtheta_0^{s+1}}
		\vert\tilde{\vtheta}^s},
\end{align*}
where the sequence $\tilde{\vtheta}^s,\vtheta_0^{s+1},\dots,\vtheta_t^{s+1}$ is defined in Algorithm \ref{alg:svrpg}, ${\it {\bf N}} = [0,\dots,N)$, and ${\it {\bf B}} = [0,\dots,B)$. 
\end{definition}
Intuitively, the $\Ets{\cdot}$ operator computes the expected value with respect to the sampling of trajectories from the snapshot $\tilde{\vtheta}^s$ up to the $t$-th iteration included. Note that the order in which expected values are taken is important since each $\vtheta_{t}^{s+1}$ is function of previously sampled trajectories and is used to sample new ones.
